MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-07-28 latex/paper/sections/10-placement.tex
cooling schedules
start talking about ap

The state \(s\) of the system and the function \(E(s)\) to be minimized is analogous to the internal energy of the system. 
In metallurgic Annealing, the state \(s\) represents the position of every atom in the metal object at any given snapshot in time, while \(E(s)\) represents the defectiveness in the metal's structure however that may be quantified. 
In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
We will be measuring wirelength using the manhattan distance between the 2-dimensional coordinates of the \texttt{SiteInst}s on the \texttt{device}.
We will refer to these distances as half perimeter wirelength (HPWL).
The goal of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.

@@

\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance}, label={lst:sa_acceptance}]
protected List<Double> coolingSchedule;
public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) {
    double currentTemp = initialTemp;
    for (int i = 0; i < movesLimit; i++) {
        this.coolingSchedule.add(currentTemp);
        currentTemp *= alpha; // geometric cooling
    }
}

@@

\subsection{Variations}
\label{subsec:variations}

@@

In force directed placers, there is a concept of global placement followed by detailed placement.
Global placement places object instances freely on an analog plane and is driven purely by directed movement and analytical solution.
In global placement stage, only non-hardware related objectives are considered. 
Object instance placements do not have to adhere to any hardware constraints and can even overlap with one another.
Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the instances onto the discrete grid of resources of the \texttt{device}.
Overlapping instances are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy. 
There also exists other strategies to mix packing, placement, and legalization in different orders. 
In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

>>>>

The state \(s\) of the system and the function \(E(s)\) to be minimized is analogous to the internal energy of the system. 
In metallurgic Annealing, the state \(s\) represents the position of every atom in the metal object at any given snapshot in time, while \(E(s)\) represents the defectiveness in the metal's structure however that may be quantified. 
In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
We will be measuring wirelength using the manhattan distance between the 2-dimensional coordinates of the \texttt{SiteInst}s on the \texttt{device}.
We will refer to these distances as half perimeter wirelength (HPWL).
The purpose of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.

@@

\newcolumn
\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance}, label={lst:sa_acceptance}]
protected List<Double> coolingSchedule;
public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) {
    double currentTemp = initialTemp;
    for (int i = 0; i < movesLimit; i++) {
        this.coolingSchedule.add(currentTemp);
        currentTemp *= alpha; // geometric cooling
    }
}

@@

@@

In force directed placers, there is a concept of global placement followed by detailed placement.
Global placement places object instances freely on an analog plane and is driven purely by directed movement and analytical solution.
In global placement stage, only non-hardware related objectives are considered. 
Object instance placements do not have to adhere to any hardware constraints and can even overlap with one another.

Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the instances onto the discrete grid of resources of the \texttt{device}.
Overlapping instances are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy. 
There also exists other strategies to mix packing, placement, and legalization in different orders, such as those studied in \cite{ExplicitPacking}. 
In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

### 2025-07-28 latex/paper/sections/11-placement_results.tex
hybrid runs
start talking about ap

Shown in figure \ref{fig:placers_overlay} are the total HPWL curves over number of passes for each of the five placers. 
\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590K.
This is likely due to the lack of randomness or lack of diversity in move proposals combined with the persistent greediness of the movement acceptance that prevents the system from escaping local minima.
This conclusion is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600K mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.

@@

\begin{multicols}{2}
    In addition to variations on the basic SA algorithm, we also test various cooling schedules on the basic SA algorithm.
    Shown in figure \ref{fig:VaryCoolingSchedules} are the cost histories for various cooling schedules, grouped into three different initial temperatures \texttt{(5000, 10000, and 15000)}. 
    For each initial temperature group, we test five different cooling rates \texttt{(0.79, 0.84, 0.89, 0.94, and 0.99)}, with the darkness of each curve corresponding with a faster cooling rate.
    The curves suggest that a lower initial temperature and faster cooling rate both contribute to a faster convergence to the global minimum, with generally small differences of less than 10\% in final HPWL cost.

>>>>

Shown in figure \ref{fig:placers_overlay} are the HPWL curves for each of the five placers. 
\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590K.
This is likely due to the lack of randomness or lack of diversity in move proposals combined with the persistent greediness of the movement acceptance that prevents the system from escaping local minima.
This conclusion is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600K mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.

@@

\begin{multicols}{2}
    In addition to variations on the basic SA algorithm, we also test various cooling schedules on the basic SA algorithm.
    Shown in figure \ref{fig:VaryCoolingSchedules} are the cost histories for various cooling schedules, grouped into three different initial temperatures \texttt{(5000, 10000, and 15000)}. 
    For each initial temperature group, we test five different cooling rates \texttt{(0.79, 0.84, 0.89, 0.94, and 0.99)}, with the darkness of each curve corresponding with a faster cooling rate.

### 2025-07-28 latex/paper/sections/12-future_work.tex
start talking about ap

In its current state, the prepacker and packer struggle to handle signals larger than 24-bits, especially when involved in DSP functions like addition and multiplication. 
In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations that are currently not handled by our packer and eventually to failures in the subsequent routing stage.
Further work is required to resolve these constraints in the packer.

\subsection{Variations on Packing}
Our current placer follows a \texttt{Site}-centric approach that resembles that of Xilinx ISE, which is the predecessor to the Vivado design suite that was primarily used in the 90s and 2000s.
The more modern Vivado now performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 
Enabling BEL-centric movement as opposed to Site-centric movement in our placer can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.

There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow. 
Future work can 
{
    \centering
    \includegraphics[width=\columnwidth]{figures/future_work/legalization.png}
    \captionof{figure}{Representative FPGA placement and packing flows. Figure taken from Wuxi et al. (2019), page 1 \cite{ExplicitPacking}}
}
\vspace{0.25cm}

@@

\subsection{Improvements to SA}

\subsection{Force-Directed and Analytical Placement}

>>>>

In its current state, the prepacker and packer struggle to handle signals larger than 24-bits, especially when involved in DSP functions like addition and multiplication. 
In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations which are currently not handled by our packer and lead to failures in the subsequent routing stage.
Further work is required to resolve these constraints in the packer.

\subsection{Force-Directed and Analytical Placement}
As mentioned previously, SA has largely been abandoned in favor of AP in SOTA placers, owing to SA's poor scalability and runtime.
Our placer follows a basic prepack-pack-place flow, where the final placement stage is driven by simulated annealing. 
In future work, we can retrofit this flow by replacing SA with AP in the final placement stage while fully reusing the prepacker and packer in its current state. 
We can still treat \texttt{SiteInst}s as the atomic objects to be placed and use analytical solvers to minimize the total HPWL of the nets between them.

The following papers can serve as foundational references for building an AP placer or insight into the SOTA placers.
\begin{itemize}
    \item \emph{Analytical minimization of half-permieter wirelength} - \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
    \item \emph{SimPL: an algorithm for placing VLSI circuits} - \textbf{Kim et al. (2013)} \cite{SimPL}.
    \item \emph{Analytical placement for heterogeneous FPGAs} - \textbf{Gort et al. (2012)} \cite{AP_2012}.
    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL-SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} - \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macros Groups and Fence Regions} - \textbf{Jing et al. (2024)} \cite{OpenPARF}.
\end{itemize}

\subsection{Variations on Packing}
Our current placer follows a \textbf{Site-centric} approach that resembles that of the old Xilinx ISE - the predecessor to Vivado used in the 90s and 2000s.
These days, Vivado performs \textbf{BEL-centric} placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 
Enabling BEL-centric moveme
nt as opposed to Site-centric movement can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.

There is no constraint that forces us to pack \texttt{Cells} into \texttt{SiteInst}s before placement or to follow a strict prepacking-packing-placement flow. 
After implementing a force-directed or analytical placer, we can begin to explore different packing and placement ordering like those studied in Wuxi et al. (2019) \cite{ExplicitPacking}.

\begin{equation}
    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i, j} w_{i, j} ( |x_i - x_j| + |y_i - y_j| )
\end{equation}

{
    \centering
    \includegraphics[width=\columnwidth]{figures/future_work/legalization.png}
    \captionof{figure}{Representative FPGA placement and packing flows. Figure taken from Wuxi et al. (2019), page 1 \cite{ExplicitPacking}}
}
\vspace{0.25cm}

@@
